---
title: "Cyclistic Analysis"
author: "Rachel Morton"
date: "2023-09-28"
output: html_document
---


## This analysis is a case study from the Google Data Analytics certificate. It is originally based on a case study written by Kevin Hartman, which can be found here: https://artscience.blog/home/divvy-dataviz-case-study

## The purpose of this analysis is to aggregate multiple datasets into a single data frame, then conduct a simple analysis to answer the business question "In what ways do members and casual riders use bike share services differently?" for our stakeholders, Cyclistic Marketing Analytics team, and Executive team.

## The datasets used in this analysis are from 2016, broken down by quarter. They were sourced from Motivate International Inc. under this license: https://divvybikes.com/data-license-agreement. After looking through a few different years of data available, I opted to work with 2016 as it seemed to be relatively unbiased, valid, and accessible. This data should help us identify differences in ridership between members and casual users.

## To see the non technical report in it's entirety, please follow the link: https://www.canva.com/design/DAFwPSINc7M/2f3e9C4vT6Tsvw7Y7Bg99g/edit

## To find this project on Kaggle, please see kaggle link in bio or follow this link: https://www.kaggle.com/code/rachelmorton/cyclistic-market-analysis



# Step 1: Prepare Environment for Analysis
#### Loading the packages necessary to complete the analysis and setting the working directory to simplify calls to data

```{r library and directory, warning=FALSE}
# Make sure to install packages before loading if you haven't already.
library(tidyverse) #- for wrangling data
library(lubridate) #- for date functions
library(dplyr) #- to manipulate data
library(readxl) #- to import data
library(scales) #- to customize plot formatting
setwd("C:/Users/Rachel/Documents/Merit America/Portfolio Projects/coursera capstone/cyclist") #Make sure to input your own directory path here!
getwd()

```



# Step 2: Import and Standardize Data
#### Importing the datasets

```{r import data}
q1_2016 <- read_xlsx("Divvy_Trips_2016_Q1_xlsx.xlsx")
q2.1_2016 <- read_xlsx("Divvy_Trips_2016_Q2.1_xlsx.xlsx")
q2.2_2016 <- read_xlsx("Divvy_Trips_2016_Q2.2_xlsx.xlsx")
q3.1_2016 <- read_xlsx("Divvy_Trips_2016_Q3.1_xlsx.xlsx")
q3.2_2016 <- read_xlsx("Divvy_Trips_2016_Q3.2_xlsx.xlsx")
q4_2016 <- read_xlsx("Divvy_Trips_2016_Q4_xlsx.xlsx")

```

#### Checking column names for consistency

```{r colnames, warning=FALSE}
colnames(q1_2016)
colnames(q2.1_2016)
colnames(q2.2_2016)
colnames(q3.1_2016)
colnames(q3.2_2016)
colnames(q4_2016)

```

#### Standardizing column names in preparation for aggregation

```{r rename columns, warning=FALSE}
(q1_2016 <- rename(q1_2016
                   ,trip_id=trip_id
                   ,bike_id=bikeid
                   ,start_time=starttime
                   ,stop_time=stoptime
                   ,trip_duration_seconds=tripduration
                   ,from_station_id=from_station_id
                   ,to_station_id=to_station_id
                   ,user_type=usertype
                   ,from_station_name=from_station_name
                   ,to_station_name=to_station_name
                   ,gender=gender
                   ,birth_year=birthyear
                   ,ride_length=ride_length
                   ,day_of_week=day_of_week))

(q2.1_2016 <- rename(q2.1_2016
                   ,trip_id=trip_id
                   ,bike_id=bikeid
                   ,start_time=starttime
                   ,stop_time=stoptime
                   ,trip_duration_seconds=tripduration
                   ,from_station_id=from_station_id
                   ,to_station_id=to_station_id
                   ,user_type=usertype
                   ,from_station_name=from_station_name
                   ,to_station_name=to_station_name
                   ,gender=gender
                   ,birth_year=birthyear
                   ,ride_length=ride_length
                   ,day_of_week=day_of_week))

(q2.2_2016 <- rename(q2.2_2016
                   ,trip_id=trip_id
                   ,bike_id=bikeid
                   ,start_time=starttime
                   ,stop_time=stoptime
                   ,trip_duration_seconds=tripduration
                   ,from_station_id=from_station_id
                   ,to_station_id=to_station_id
                   ,user_type=usertype
                   ,from_station_name=from_station_name
                   ,to_station_name=to_station_name
                   ,gender=gender
                   ,birth_year=birthyear
                   ,ride_length=ride_length
                   ,day_of_week=day_of_week))

(q3.1_2016 <- rename(q3.1_2016
                   ,trip_id=trip_id
                   ,bike_id=bikeid
                   ,start_time=starttime
                   ,stop_time=stoptime
                   ,trip_duration_seconds=tripduration
                   ,from_station_id=from_station_id
                   ,to_station_id=to_station_id
                   ,user_type=usertype
                   ,from_station_name=from_station_name
                   ,to_station_name=to_station_name
                   ,gender=gender
                   ,birth_year=birthyear
                   ,ride_length=ride_length
                   ,day_of_week=day_of_week))

(q3.2_2016 <- rename(q3.2_2016
                   ,trip_id=trip_id
                   ,bike_id=bikeid
                   ,start_time=starttime
                   ,stop_time=stoptime
                   ,trip_duration_seconds=tripduration
                   ,from_station_id=from_station_id
                   ,to_station_id=to_station_id
                   ,user_type=usertype
                   ,from_station_name=from_station_name
                   ,to_station_name=to_station_name
                   ,gender=gender
                   ,birth_year=birthyear
                   ,ride_length=ride_length
                   ,day_of_week=day_of_week))

(q4_2016 <- rename(q4_2016
                   ,trip_id=trip_id
                   ,bike_id=bikeid
                   ,start_time=starttime
                   ,stop_time=stoptime
                   ,trip_duration_seconds=tripduration
                   ,from_station_id=from_station_id
                   ,to_station_id=to_station_id
                   ,user_type=usertype
                   ,from_station_name=from_station_name
                   ,to_station_name=to_station_name
                   ,gender=gender
                   ,birth_year=birthyear
                   ,ride_length=ride_length
                   ,day_of_week=days_of_week))
```

#### Verifying consistency of column names and data types

```{r inspecting dataframes, echo=TRUE, warning=FALSE}
str(q1_2016)
str(q2.1_2016)
str(q2.2_2016)
str(q3.1_2016)
str(q3.2_2016)
str(q4_2016)

```



# Step 3: Aggregate Data for Analysis
#### Aggregating all of the datasets into one data frame for analysis

```{r aggregate dataframes, warning=FALSE}
all_trips <- bind_rows(q1_2016, q2.1_2016, q2.2_2016, q3.1_2016,q3.2_2016, q4_2016)

```

#### Creating a secondary single data frame to manipulate data as a save point in case of error

```{r creating secondary aggregated data frame before editing data, warning=FALSE}
all_trips_edited <-all_trips

```



# Step 4: Clean Data for Analysis
#### Remove extra data

```{r remove extra data, echo=TRUE, warning=FALSE}
all_trips_edited <- all_trips_edited %>%
  select(-c(birth_year, gender, trip_duration_seconds, ride_length))
# Removed ride_length due to errors with the data later in the script, added back in as calculated field
```

#### Inspecting the new data frame 

```{r inspect new dataframe, echo=TRUE, warning=FALSE}
colnames(all_trips_edited)
nrow(all_trips_edited)
dim(all_trips_edited)
head(all_trips_edited)
str(all_trips_edited)
summary(all_trips_edited)

```

#### Updating user type column to current values used by Cyclistic

```{r update user_types, echo=TRUE, warning=FALSE}
all_trips_edited <- all_trips_edited %>%
  mutate(user_type=recode(user_type
                ,"Subscriber"="member"
                ,"Customer"="casual"))

table(all_trips_edited$user_type) #checking the output
```

#### Adding columns to break down start time into pieces

```{r add column for date elements at start of trip, echo=TRUE, warning=FALSE}
all_trips_edited$date <- as.Date(all_trips_edited$start_time)
all_trips_edited$month <- format(as.Date(all_trips_edited$date), "%m")
all_trips_edited$day<- format(as.Date(all_trips_edited$date), "%d")
all_trips_edited$year<- format(as.Date(all_trips_edited$date), "%Y")
all_trips_edited$weekday<- format(as.Date(all_trips_edited$date), "%A")

```

#### Adding a calculated field for ride length

```{r calculating ride_length, echo=TRUE, warning=FALSE}
all_trips_edited$ride_length <- difftime(all_trips_edited$stop_time, all_trips_edited$start_time, units = "mins")

```

#### Inspecting the data frame structure

```{r inspect the structure of the columns, echo=TRUE, warning=FALSE}
str(all_trips_edited)

```

#### Converting ride length to a numeric datatype

```{r convert ride_length, warning=FALSE}
is.factor(all_trips_edited$ride_length)
all_trips_edited$ride_length <- as.numeric(as.character(all_trips_edited$ride_length))
is.numeric(all_trips_edited$ride_length)

```

#### Removing bad data in user types

```{r removing bad user_types, warning=FALSE}
all_trips_edited <- all_trips_edited[!(all_trips_edited$user_type == "Dependent" | all_trips_edited$user_type<0),]

```



# Step 5: Analyze the Data
#### Analyzing ride length

```{r descriptive analysis of ride_length, echo=TRUE, warning=FALSE}
mean(all_trips_edited$ride_length)
median(all_trips_edited$ride_length)
min(all_trips_edited$ride_length)
max(all_trips_edited$ride_length)

```

#### Comparing members vs. users

```{r compare members vs casual users, echo=TRUE, warning=FALSE}
aggregate(all_trips_edited$ride_length ~ all_trips_edited$user_type, FUN = mean)
aggregate(all_trips_edited$ride_length ~ all_trips_edited$user_type, FUN = median)
aggregate(all_trips_edited$ride_length ~ all_trips_edited$user_type, FUN = min)
aggregate(all_trips_edited$ride_length ~ all_trips_edited$user_type, FUN = max)

```

#### Calculating average ride time between user types

```{r avg. ride time between user types per day, echo=TRUE, warning=FALSE}
aggregate(all_trips_edited$ride_length ~ all_trips_edited$user_type + all_trips_edited$weekday, FUN = mean)

aggregate(all_trips_edited$ride_length ~ all_trips_edited$user_type + all_trips_edited$month, FUN = mean)

avg_ride_length_week <- aggregate(all_trips_edited$ride_length ~ all_trips_edited$user_type + all_trips_edited$weekday, FUN = mean)
avg_ride_length_month <- aggregate(all_trips_edited$ride_length ~ all_trips_edited$user_type + all_trips_edited$month, FUN = mean)
```

#### Putting the weekdays in order

```{r correct weekday order in previous chunk output, warning=FALSE}
all_trips_edited$weekday <- ordered(all_trips_edited$weekday, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

```

#### verifying output

```{r avg. ride time between user types per day ordered, echo=TRUE, warning=FALSE}
aggregate(all_trips_edited$ride_length ~ all_trips_edited$user_type + all_trips_edited$weekday, FUN = mean)

```

#### Calculating ridership by user type and time period

```{r ridership by type and weekday, echo=TRUE, warning=FALSE}
all_trips_edited %>%
  mutate(weekday=wday(start_time, label= TRUE)) %>%
  group_by(user_type, weekday) %>%
  summarize(number_of_rides=n()
            , average_duration = mean(ride_length)) %>%
            arrange(user_type, weekday)
            
```
```{r ridership by type and month, echo=TRUE, warning=FALSE}
all_trips_edited %>%
  mutate(weekday=wday(start_time, label= TRUE)) %>%
  group_by(user_type, month) %>%
  summarize(number_of_rides=n()
            , average_duration = mean(ride_length)) %>%
            arrange(user_type, month)


```


#### Saving Tibbles as data frames for use in viz


```{r turning that tibble into a data frame for later use}

summary_data_weekday <- all_trips_edited %>%
  mutate(weekday=wday(start_time, label= TRUE)) %>%
  group_by(user_type, weekday) %>%
  summarize(number_of_rides=n()
            , average_duration = mean(ride_length)) %>%
            arrange(user_type, weekday)


```

```{r turning that tibble into a data frame for later use}

summary_data_month <- all_trips_edited %>%
  mutate(weekday=wday(start_time, label= TRUE)) %>%
  group_by(user_type, month) %>%
  summarize(number_of_rides=n()
            , average_duration = mean(ride_length)) %>%
            arrange(user_type, month)


```


# Step 6: Visualize Data
#### Plotting the number of rides by user type

```{r visualize number of rides by user_type, echo=TRUE, warning=FALSE}
all_trips_edited %>%
  mutate(weekday=wday(start_time, label= TRUE)) %>%
  group_by(user_type, weekday) %>%
  summarise(
    'number_of_rides' = n(),
    'average_duration' = mean(ride_length, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill= user_type)) +
  geom_col(position = "dodge") +
  scale_y_continuous(
    labels = c(0, 10000, 20000, 30000, 40000, 50000))

```




#### Plotting the number of rides by average duration of ride length

```{r visualize average_duration of ride_length, echo=TRUE, warning=FALSE}
all_trips_edited %>%
  mutate(weekday=wday(start_time, label= TRUE)) %>%
  group_by(user_type, weekday) %>%
  summarise(
    'number_of_rides' = n(),
    'average_duration' = mean(ride_length, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = weekday, y = average_duration, fill= user_type)) +
  geom_col(position = "dodge")

```



# Step 7: Export Summary File for Further Analysis
#### Exporting a summary data frame for visualization and presentation

```{r export summary data, warning=FALSE}
write.csv(summary_data_weekday, file ="C:/Users/.../summary_weekly.csv")
write.csv(summary_data_month, file ="C:/Users/.../summary_monthly.csv")
write.csv(all_trips_edited, file ="C:/Users/.../all_trips.csv")
```
